---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center"
)
```

# ggdemetra

[![Build Status](https://api.travis-ci.com/AQLT/ggdemetra.svg?branch=master)](https://travis-ci.com/AQLT/ggdemetra)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggdemetra)](https://cran.r-project.org/package=ggdemetra)
[![CRAN last release](http://www.r-pkg.org/badges/last-release/ggdemetra)](https://cran.r-project.org/package=ggdemetra)
[![CRAN monthly downloads](http://cranlogs.r-pkg.org/badges/ggdemetra?color=lightgrey)](https://cran.r-project.org/package=ggdemetra)
[![CRAN downloads](http://cranlogs.r-pkg.org/badges/grand-total/ggdemetra?color=lightgrey)](https://cran.r-project.org/package=ggdemetra)

## Overview
ggdemetra is an extension of [ggplot2](https://github.com/tidyverse/ggplot2) to add seasonal adjustment statistics to your plots. The seasonal adjustment process is done with [RJDemetra](https://github.com/jdemetra/rjdemetra) that is an R interface to [JDemetra+](https://github.com/jdemetra/jdemetra-app), the seasonal adjustment software [officially recommended](https://ec.europa.eu/eurostat/cros/system/files/Jdemetra_%20release.pdf) to the members of the European Statistical System (ESS) and the European System of Central Banks. RJDemetra implements the two leading seasonal adjustment methods [TRAMO/SEATS+](http://www.bde.es/bde/en/secciones/servicios/Profesionales/Programas_estadi/Programas_estad_d9fa7f3710fd821.html) and [X-12ARIMA/X-13ARIMA-SEATS](https://www.census.gov/srd/www/x13as/).

There are 4 main functionnalities in `ggdemetra` depending of what you want to add in the graphic:

- `geom_sa()`: to add a time series compute during the seasonal adjustment (the trend, the seasonal adjusted time series, etc.).  
- `geom_outliers()`: to add the outliers used in the pre-adjustment process of the seasonal adjustment.
- `geom_arima()`: to add the ARIMA model  used in the pre-adjustment process of the seasonal adjustment.
- `geom_diagnostics()`: to add a table containing some diagnostics on the seasonal adjustment process.

## Installation

Since RJDemetra requires Java SE 8 or later version, the same requirements are also needed for ggdemetra.

```{r, eval = FALSE}
# Install release version from CRAN
install.packages("ggdemetra")

# Install development version from GitHub
# install.packages("devtools")
devtools::install_github("AQLT/ggdemetra")
```

## Usage

To add the seasonal adjusted series and the forecasts of the input data and of the seasonal adjusted series:
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggdemetra)

p_ipi_fr <- ggplot(data = ipi_c_eu_df, mapping = aes(x = date, y = FR)) +
    geom_line() +
    labs(title = "Seasonal adjustment of the French industrial production index",
         x = "time", y = NULL)
p_sa <- p_ipi_fr +
    geom_sa(component = "y_f", linetype = 2, message = TRUE) + 
    geom_sa(component = "sa", color = "red", message = FALSE) +
    geom_sa(component = "sa_f", color = "red", linetype = 2, message = FALSE)
p_sa
```

To add the outliers at the bottom of the plot with an arrow to the data point and the estimate coefficient:
```{r}
p_sa + 
    geom_outlier(geom = "label_repel",
                 coefficients = TRUE,
                 message = FALSE,
                 vjust = 4,
                 ylim = c(NA, 65), force = 10,
                 arrow = arrow(length = unit(0.03, "npc"),
                               type = "closed", ends = "last"))
```

To add the ARIMA model:

```{r}
p_sa + 
    geom_arima(geom = "label",
               x_arima = -Inf, y_arima = -Inf, 
               vjust = -1, hjust = -0.1,
               message = FALSE)
```

To add a table of diagnostics below the plot:
```{r}
diagnostics <- c(`Combined test` = "diagnostics.combined.all.summary",
                 `Residual qs-test (p-value)` = "diagnostics.qs",
                 `Residual f-test (p-value)` = "diagnostics.ftest")
p_diag <- ggplot(data = ipi_c_eu_df, mapping = aes(x = date, y = FR)) +
    geom_diagnostics(diagnostics = diagnostics,
                     table_theme = gridExtra::ttheme_default(base_size = 8),
                     message = FALSE) + 
    theme_void()
    
gridExtra::grid.arrange(p_sa, p_diag,
             nrow = 2, heights  = c(4, 1))
```
